Exploiting First-Class Arrays in Fortran for Accelerator Programming

### 2010-12-10 opencl/papers/foropencl.tex
Added a lot of text from proposal and the book.  Still need to weed it down.

@@

At Los Alamos we have a wide variety of computer platforms from which
to choose.  The common platform is made up of cluster of compute nodes
with standard multi-core processors.  Some nodes have accelerators
that range from the IBM Cell processor (Roadrunner) to a variety of
GPUs from NVIDIA and AMD.  Some nodes have have hardware with vector
instructions and others do not.  An important question is what programming
model and what languages are best suited for these platforms.

@@

This papers addresses this issue by examining features in Fortran that
allow programmers to express algorithms at a very high level but yet
can be transformed by a compiler to run efficiently on a wide variety
of platforms.  In particular we consider nodes of GPUs.

@@

We study automatic transformations of Fortran for two applications examples.
The first is a simple shallow water model in two dimensions using
finite volume methods with .... for time update.  The second is based
on the primary compute kernels from PetaVision, a high-performance, 
neural science framework that is used in models of visual cortex.

@@

@@

@@

@@

This paper uses four simple features of Fortran that form a language subset
that does not require any language extensions and can be easily transformed
to a lower level implementation in either CUDA or OpenCL.

@@

@@

@@

The final piece that we exploit is the array notation present in modern
versions of Fortran.  ....

>>>>

%
% I'm adding an outline here so I can see how everything fits together

%
% \title{Exploiting First-Class Arrays in Fortran for Accelerator Programming}
%

%
% \abstract
%

%
% \section{Introduction}
%    \subsection{Approach}
%    \subsection{Why Fortran?}
%    \subsection{Other Languages}
%

%
% \section{Programming model}
%    \subsection{Fortran Syntax}   
%       \subsubsection*{Array notation}
%       \subsubsection*{Elemental functions}
%       \subsubsection*{Pure procedures}
%       \subsubsection*{Shift operators}
%       \subsubsection*{Regions}
%    \subsection{Parallelism}
%    \subsection{New Functions}
%    \subsection{Limitations}
%

%
% \section{Source-To-Source Transformations}
%    \subsection{ForOpenCL}
%       \subsubsection{array syntax}
%       \subsubsection{where construct}
%    \subsection{New functions}
%       \subsubsection{region}
%    \subsection{Static Analysis}
%       \subsubsection{Analysis not required}
%    \subsection{Simplifying Assumptions}
%

%
% \section{Shallow Water Model}
%    \subsection{Equations}
%

%
% \section{Performance}
%

%
% \section{Conclusions}
%

@@

At Los Alamos National Laboratory (LANL) we have a wide variety of computer
platforms from which to choose.  The common platform is made up of clusters of
compute nodes with standard multi-core processors.  Some nodes also have
accelerators that range from the IBM Cell processor (Roadrunner) to a variety
of GPUs from NVIDIA and AMD.  Some nodes have hardware with vector
instructions and others do not. The peak
performance of the accelerated nodes range from 200+ GFlops for the IBM Cell/B.E.
processor to XXX for NVIDIA Fermi. %% Fermi has another name

%% NEW - CER
However, the performance that these accelerator architectures offer comes at a
cost as processor architectures are trending toward multiple cores with
instances of integrated accelerator units and less of a reliance on memory
caches.  These changes place a heavy burden on application programmers as they
are forced to adapt to the new systems.  An especially challenging problem
faced by application programmers is not only how to program to these new
architectures (considering the massive scale of concurrency available), but
also how to design programs that are portable across the changing landscape of
computer architectures.  The fundamental question we address is what
programming model and what languages are best suited for these platforms.

@@

This paper addresses this issue by examining features in Fortran that
allow programmers to express algorithms at a very high level but yet
can be transformed by a compiler to run efficiently on a wide variety
of platforms.  In particular we consider nodes of GPUs.

@@

%% NEW - CER
The sheer complexity of programming for clusters of many or multi-core
processors with tens of millions threads of execution make the simplicity of
the data parallel model attractive.  Furthermore, the increasing complexity of
todays applications (especially when convolved with the increasing complexity
of the hardware) and the need for portability across hardware architectures
make a higher-level and simpler programming model like data parallel
attractive.

%% NEW - CER
The goal of this work is to exploit source-to-source transformations that
allow programmers to develop and maintain programs at a high-level of
abstraction, without coding to a specific hardware architecture.
Furthermore these transformations allow multiple hardware architectures
to be targeted without changing the high-level source.  It also removes the
necessity for application programmers to understand details of the accelerator
architecture or to know OpenCL.

We use the ROSE compiler infrastructure \cite{} to develop these
transformations.  ROSE uses the Open Fortran Parser \cite{} to parse Fortran
2008 syntax and can output (unparse) to OpenCL.  Since ROSE's intermediate
representation (IR) was constructed to represent multiple languages, it is
relatively straightforward to transform high-level Fortran IR nodes to OpenCL
nodes.  Furthermore, the Fortran array syntax maps directly to one, two, and
three-dimensional thread groups in OpenCL.

%%We study automatic transformations of Fortran for two applications examples.  The first is a simple shallow water model in two dimensions using finite volume methods with .... for time update.  The second is based on the primary compute kernels from PetaVision, a high-performance, neural science framework that is used in models of visual cortex.

%% To simplify things, lets save PetaVision for the local stuff as it requires convolutions

We study automatic transformations of Fortran for an application example that
is typical of many applications that are based on finite-difference or
finite-volume methods in computational fluid dynamics (CFD).  The example is a
simple shallow water model in two dimensions using finite volume methods with
.... for time update.

%% NEW - CER
PAGOSA is a production code at LANL completely written in data-parallel
Fortran.  In a prior work we investigated automatically transforming this code
to run on LANL's Petaflop Roadrunner computer (a hybrid mixture of Opterons
and IBM Cell processors).  We demonstrated that a source-to-source compiler
can automatically vectorize and parallelize (both within the Cell processor
and across nodes) a small section of this code.  Preliminary results showed a
12 times performance gain on the Cell processor over a traditional single core
processor.  A factor of 9 is gained from vectorizing and parallelizing the
code to run on the Cell and the rest is gained from automatically overlapping
inter-node communication with computation on the Cell.

@@

%% NEW - CER
Fortran \footnote{Originally named FORTRAN for FORmula TRANslation language.}
is the oldest high-level programming language, and was developed to facilitate
the translation of math formulae into machine code. Fortran was the first
major language to use a compiler to translate from a high-level program
representation to assembly language. Due to its age, it carries certain arcane
baggage that later programming languages have evolved away from. As a result,
Fortran has fallen into disfavor in certain programming circles.

%% NEW - CER
However, with the introduction of Fortran 90 (and later revisions to the
standard), Fortran is a truly modern programming language. Much of the arcane
baggage associated with it is actually no longer part of the language used by
modern programmers, and exists in a deprecated form simply to support legacy
codes. It is now modular and has many object-oriented features. Furthermore,
Fortran should be of interest to those studying parallel programming because
of its functional and data parallel constructs and because of the coarray
notation introduced after Fortran 2003. Unlike languages like C and C++, which
are considered to be more modern, Fortran has become a truly parallel language
with features added to recent language standards. With coarrays, Fortran
provides a mechanism to run replicas (called images) of a program on many
different processors (or cores) at the same time. It also provides language
constructs that allow the program to both read from and write to data residing
on {\it remote} processors and to synchronize the different program images so
that access to the data by the multiple images can be done atomically.

@@

%% NEW - CER
So perhaps it is time to replace Fortran with yet another computer language.  The
problems with replacing Fortran with an entirely new language are two fold:
the economics of replacing the existing application base and the difficulty in
obtaining programmer acceptance.  It is estimated that replacing a major
production application at Los Alamos National Laboratory would cost between 50
and 150 million dollars.  In terms of programmer acceptance, there is always
the "chicken and egg problem": programmers won't use a new language until they
can expect good performance across a variety of platforms, and compiler
vendors can't afford to produce quality compilers until there is a reasonable
expectation of a market.

@@

%% NEW - CER
History has also shown that an investment in rewriting code does not guarantee
success either, as seen in an effort at LANL to modernize a legacy Fortran
code with the the newer C++ POOMA framework.  This massive overhaul effort led
to a code that was both slower and less flexible than the original Fortran
\cite{}.  Similar experiences have occurred in the past, notably during the
devlopment of the Sisal language.

@@

%% NEW - CER
We will examine how the existing data-parallel constructs in Fortran can be
combined with coarrays or MPI to provide effectively a new parallel programming
language, one that is evolutionary in nature and provides complete compatibility
with existing applications and libraries.  Data parallelism is a high-level
abstraction that is, at the same time, both easier to program and gives the
compiler more leeway (if fully exploited) in retargeting a program to different
computer architectures.

%% NEW - CER
An example of this style of programming in Fortran is shown in

\begin{verbatim}

  Bz = Bz + dt * (cshift(Ex,dim=2,shift=+1) - Ex) / dy  &
          - dt * (cshift(Ey,dim=1,shift=+1) - Ey) / dx

\end{verbatim}

%% NEW - CER
This example is a solution to Maxwell's equations for the $z$ component of the
magnetic field using Fortran array syntax.  Note that there are no explicit
loops in this example.  The operators {\tt +}, {\tt -}, and {\tt *} are
applied to all of the elements of the three-dimensional arrays {\tt Bz}, {\tt
  Ex}, and {\tt Ey}, individually.  This is why data parallelism has been
called collection-oriented programming by
Blelloch~\cite{blelloch90,rajopadhyedidlacs}.  As the {\tt cshift} function
and the array-valued expressions all semantically returns a value, this style
of programming is also similar to functional programming (or value-oriented
programming). %%~\cite{simonpeytonjones}).  The heart of the solution to
Maxwell's equations is the statement shown in Listing~\ref{lst:dpexample} and
five similar, simple equations.

%% NEW - CER
The power in this notation is that the compiler is free to distribute the
computation in these expressions over any hardware processing elements available
to it, such as the vector units on an Intel or AMD processor or the Synergistic
Processing Units on an IBM Cell processor.  If the arrays are declared as
coarrays, this includes spreading the computation over \emph{all of the nodes in
  a cluster as well.}  In this case, communication occurs within the
\texttt{cshift} functions, though the compiler is free to overlap communication
and computation by scheduling the communication early and performing the
computation on the interior of the arrays while waiting for the communication to
complete.

%% NEW - CER
Complete and very concise and elegant programs can be built with statements
similar to that shown in Listing~\ref{lst:dpexample}, using intrinsic
functions like the array constructors (\texttt{CSHIFT}, \texttt{MERGE},
\texttt{TRANSPOSE}, ...), the array location functions (\texttt{MAXLOC} and
\texttt{MINLOC}), and the array reduction functions (\texttt{ANY},
\texttt{COUNT}, \texttt{MINVAL}, \texttt{SUM}, \texttt{PRODUCT},
...)~\cite{guydp}.  It should be emphasized, however, that {\it no} compiler
currently supplies coarray versions of these intrinsic functions as they are
not defined by the proposed Fortran 2008 standard (we have submitted a paper
proposing the integration of coarrays with these intricics functions to the J3
Fortran committee~\cite{rasmussen-caf-intrinsics}).  With coarray or MPI
implementations of these intrinsic functions, programs written in this
data-parallel subset of Fortran are implicitly parallel.  Furthermore, in
instances it is necessary to implement algorithms that don't lend themselves
to expressibility in this data-parallel subset, coarray notation can be used
for complete access to individual (including remote) data elements.  It is the
combination of data parallel and coarray constructs, with the transformational
properties of a source-to-source compiler (transforming array notation to
optimized Fortran 95 and library calls), that make this possibility such a
promising avenue for research.

%% NEW - CER
This style of programming meets the requirements we have set for a programming
model for developing petascale applications.  It allows the programmer to
program at a very-high level of abstraction while providing the compiler with
maximum flexibility in targeting the application for a particular hardware
architecture.  The data parallel programming model simultaneously meets the
seemingly conflicting goals of maintainability, portability, and performance.

%% NEW - CER
Unfortunately, this style of programming has never really caught on because
when Fortran 90 was first introduced, performance was relatively poor and thus
programmers shied away from using array syntax (even now, some are actively
counseling against its usage because of performance issues~\cite{Cray-SC07}).
Thus the Fortran community was caught in a classic ``chicken-and-egg''
conundrum: (1) programmers didn't use it because it was slow; and (2)
compilers vendors didn't improve it because programmers didn't use it.

%% NEW - CER
The goal of this research is to demonstrate that parallel programs written in
this style of Fortran are maintainable and can achieve good performance on a
variety of processor architectures, including possibly, future terascale
architectures from Intel (such as the Larrabee architecture).  Using a code
segment from a production code at Los Alamos (written entirely in this style
of Fortran), we have demonstrated the potential for source-to-source
transformations to target the IBM Cell architecture used in Los Alamos
National Laboratories' Roadrunner computer~\footnote{Roadrunner is the first
  computer to break the petaflop barrier.}.  Initial results show that
transformed code using OFP and ROSE can achieve a speedup of up to 13 times on
sixteen Roadrunner nodes, compared with the equivalent number of single-core
Opteron nodes.

This paper uses four simple features of Fortran that form a language subset
that does not require any language extensions and can be easily transformed
to a lower-level implementation in either CUDA or OpenCL.

@@

%% NEW - CER
The use of array notation allows one to program in a data-parallel subset of
Fortran. This style of programming makes use of array notation and pure and
elemental functions to operate on array elements.

%% NEW - CER
Because elemental functions return scalar values and are free from side
effects, the compiler is free to distribute the computation over any hardware
processing elements available to it, such as the multiple cores and vector
units on an Intel or AMD processor or the Synergistic Processing Units on an
IBM Cell processor.

%% NEW - CER
An example of an elemental function is shown in Listing~\ref{lst:elemental}.
This example simply returns its input divided by 2. While this may seem like a
trivial example, such simple functions may be composed with other elemental
functions to perform powerful computations, especially when applied to arrays.

@@

%% NEW - CER
The use of an elemental function is shown in the pure procedure {\tt laplace}
in Listing~\ref{lst:pure}. This function (for example) can be used to solve
the steady-state temperature along a one-dimensional surface. The temperature
at any point along the surface is equal to the average of the temperature at
neighboring points. Rather than writing explicit loops and using indexing to
reference neighboring points, one may simply shift the array {\tt T} to the
left and right, add the results and divide by 2. The intrinsic function {\tt
EOSHIFT} (end-off shift) semantically returns an array copy with the value at
any element shifted by the amount given by the {\tt shift} argument. The value
shifted in at the boundary is supplied by the {\tt boundary} argument. The
elemental function {\tt one\_half} is used to divide by 2.

%% NEW - CER
Note that semantically the elemental functions return their array results by
value. This is important because had loops been used to express the {\tt
laplace} procedure, temporary variables would have been required. Otherwise,
{\tt T(i)} would be updated with new values of {\tt T(i-1)}.

%% NEW - CER
 A ``pure'' function is one
without side effects (though it may change the value of array elements in an
array argument). An elemental function is a pure function with the additional
constraint that it takes only scalar arguments and must return a scalar.
Though defined in terms of scalars, surprisingly, an elemental function may
take an array as an actual argument, with the compiler applying the function
to each element of the array, and returning an array result. The plus operator
as described in the previous paragraph can be seen as an elemental function,
as the plus operator is defined in terms of scalars but can be applied to
whole arrays and can return an array result.

@@

%% NEW - CER
Arrays are first class citizens in Fortran as they are part of the language
type system rather than just a pointer to raw memory as in C. A Fortran array
contains metadata associated with it that describes the array's shape and
size. This metadata is lacking in other languages, such as C, C++, and Java.
Even in the case where multidimensional array data structures are available
(such as in C++ or Java), they exist outside the language standard, limiting
the ability of the compiler to analyze them fully. The existence of this
metadata allows the use of array notation whereby explicit indexing of arrays
is not required. For example, if {\tt A}, {\tt B}, and {\tt C} are all arrays
of the same rank and shape, then the statement {\tt C = A + B} results in the
sum of elements of {\tt A} and {\tt B} being stored in the corresponding
elements of {\tt C}. The first element of {\tt C} will contain the value of
the first element of {\tt A} added to the first element of {\tt B}.

The final piece that we exploit is the array notation present in modern
versions of Fortran.  For example,

\begin{equation}
\label{eq:dataparallel}
A = B + s * C
\end{equation}

is a array assignment statement where A, B, and C are arrays and s is a
scalar.  Note that no explicit iteration over array indices is needed and that
the individual operators, plus, times, and assignment are applied by the
compiler to individual elements of the arrays independently.  Thus the
compiler is able to spread the computation in~\ref{eq:dataparallel} across any
hardware threads under its control.

There are several advantages to this style of programming:

\begin{itemize}
	\item There are no loops or index variables to keep track of.  Off by one
index errors are a common programming mistake.
        \item The written code is closer to the algorithm, easier to understand, and is usually substantially shorter.
	\item Semantically the intrinsic functions return arrays by value.  This is usually what the algorithm requires.
	\item Because pure and elemental function are free from side effects, it is easier for a compiler to schedule the work to be done in parallel.
\end{itemize}

Complete and very concise and elegant programs can be built with procedures
similar to the {\tt laplace} example shown above. To aid this effort, Fortran
supplies intrinsic functions like the array constructors (\texttt{CSHIFT},
\texttt{EOSHIFT}, \texttt{MERGE}, \texttt{TRANSPOSE}, ...), the array location
functions (\texttt{MAXLOC} and \texttt{MINLOC}), and the array reduction
functions (\texttt{ANY}, \texttt{COUNT}, \texttt{MINVAL}, \texttt{SUM},
\texttt{PRODUCT}, ...).
